Theoretical prediction of disrupted Min oscillation in flattened \emph{Escherichia coli}

### 2015-03-30 paper/plos.tex
Daivd read-through edits to paper Tested everything on crystal.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author Summary}
  When an E. Coli cell divides it must split into two relatively equal
  volumes in order to survive.  One mechanism that aids in this
  process is a system of proteins that oscillate between cell ends and
  help the original cell divide along its center.  A recent
  experiment found that by drastically deforming these cells this
  oscillation can be disrupted.  We examine these oscillations in
  similarly deformed cells computationally, using two widely used
  simulation models, and find good agreement using one of them.  The
  other model fails in a manner far more dramatic than has been
  previously observed.  Furthermore, we find that the effect observed
  experimentally in asymmetric and flattened cells is primarily a
  result of the flattening, not the asymmetry.

@@

Previous experimental studies have shown that the MinD protein system
is capable of exhibiting oscillations in round
shapes~\cite{fange2006noise} as well as in connected three pronged
tube shapes~\cite{varma2008min}, in which the oscillations seem to
seek out the extreme poles in the cell.  One conclusion of these
studies has been that the MinD system robustly forms bipolar
oscillations regardless of variations in cellular shape.

However, Mannik \emph{et al.} have recently shown that there are
limitations to this robust capability to
oscillate~\cite{mannik2012robustness,mannik2009bacterial,mannik2010bacteria}. They
have experimentally forced \emph{E. coli} cells into microfabricated
channels of thickness less than that of the natural diameter of the
cells, in which they area able to grow and divide. Upon entering the
channels, the cells undergo a mechanical deformation in which they
both widen and lengthen within the plane of the channel.  This
deformation results in very wide (they can reach widths of over
$5\micron$) flattened cells that when viewed from the top down have
irregular and asymmetric shapes.  While these cells are still able to
divide into surprisingly equal volumes, the MinD oscillations in these
cells are irregular, both temporally and spatially. Seen with
fluorescent microscopy, the MinD maximize in multiple locations within
the cell, in a seemingly random sequence. These experiments allow for
an opportunity to test MinD simulation models against more extreme
cases than have been seen thus far.  An additional interesting
question is whether the irregularity observed by Mannik \emph{et al.}
indicates that the MinD system can become chaotic (as in
e.g. Ref.~\citenum{kuramoto1978diffusion}), or if this irregularity is
the result of a stochastic but nonchaotic process.
%
A number of models of the MinD protein system have been developed that
accurately describe its basic oscillatory nature.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
approach with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is
recruited to the membrane by MinD that is already clustered there
(following observed non-linear attachment of MinD on the cell
membrane~\cite{hu2002dynamic,shih2002division}), and is stationary
once attached.
%

\blue{5a,b,c,d}Several models
\cite{fange2006noise,bonny2013membrane,halatek2012highly} modify
Huang's original model to more accurately reflect experimental
findings, which were published after his 2003
paper~\cite{meacci2006mobility,loose2011min}. These models introduce
diffusion along the membrane that is two to three orders of magnitude
slower than that within the cytoplasm, and increase the rate of
cytoplasmic diffusion by a factor of about five.  They also increase
or assume instantaneous~\cite{bonny2013membrane} the nucleotide
exchange rate.  In addition, they revise the process of MinD
recruitment to the membrane, prohibiting MinD from attaching to
MinD-MinE complexes there, behavior that is experimentally unresolved.
The model of Bonny \emph{et al.} allows MinE to remain independently
attached to the membrane after dissasociating with
MinD~\cite{bonny2013membrane}.
%

There are only on the order of 1000 MinD proteins in a given cell,
which is significantly less than the number of gridpoints used in our
simulations, suggesting that a continuum description will break down
and stochastic variation plays a significant role.  To treat this,
there have been a number of variations of the Huang 2003 model that
stochastically simulate the similar reaction-diffusion
equations~\cite{fange2006noise, kerr2006division}.  These studies
largely confirm the results of Huang's deterministic model when
applied to the wild-type, pill-shaped phenotype.  However, the
stochastic models are slightly more successful in predicting
experimentally observed oscillations in round cell
phenotypes~\cite{fange2006noise, huang2004min}, and they enable
prediction of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

@@

Huang's simulations use total MinD and MinE concentrations of
$1,000/\micron$ and $350/\micron$, respectively, in a cylindrical cell
of radius $0.5\micron$, and in our (non-cylindrical) cells we use the
same number of proteins per unit volume.  These concentration values
are $1273\micron^{-3}$ and $446\micron^{-3}$, respectively. We have
written our own simulation platform and membrane creation tools from
scratch, instead of using existing available software used in previous
studies~\cite{bonny2013membrane,fange2006noise,halatek2012highly,kerr2006division}
(we believe there is value in reproducing results with alternative
software).  Our simulations take place within a three-dimensional
Cartesian grid that has a grid spacing of .05\micron, and our cell
shapes are specified as the zero of an analytic function which can
generate the geometry presented in this paper.  Our code is publically
available on github~\cite{git}, including scripts that generate the
simulation data used in this paper.

@@

We mean to investigate the geometric limits of the Min system
oscillations as observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}, so we have modeled the Min system
in several cell shapes and sizes.  Here we present a selection of
these, beginning with naturally occurring pill-shaped cells, followed
by a number of flattened out shapes which reflect the experiments of
Mannik \emph{et al.}, in which bacteria are confined within a thin
slits.  These slits are fabricated with a width of .25\micron, but
they are coated with a PDMS lining that the cells are able to deform,
raising uncertainty in the actual cell thickness.  Tests with pure
silicon, non-deformable slits show that the greatest thickness for
which cells are able to penetrate is
.4\micron~\cite{mannik2012robustness,mannik2009bacterial}.  We
therefore assume that the PDMS slits have been deformed to this
thickness and simulate flattened cells with a thickness of .4\micron.
Viewed from the top down the cells have the shapes described below and
viewed from the side they have at their edges a semicircular
cross-section (one may imagine the shape of a pancake).
%
In this paper we focus on four specific flattened cell shapes.  Two of
these shapes replicate those published by Mannik \emph{et al.}, and
the other two are `stadium' shapes that respectively have the same
aspect ratio, thickness, and volume as the two cell shapes
experimentally observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}.  Viewed from the top down, these
stadium shapes appear as rectangles with semi-circular end caps on the
long axis ends.

@@

Huang \emph{et al.}~\cite{huang2003dynamic} has performed a linear
stability analysis on a cylindrical model which shows an upper limit
on a steady state solution of a $2\micron$ half wavelength.  Cells
with dimensions longer than this spontaneously develop spatial
oscillations, while cells that are shorter relax into a homogeneous
steady state.  We have performed a similar linear stability analysis
on an infinite slab with a thickness equal to that of our flattened
cells ($0.40\micron$).  We first solve the model's five differential
equations for the steady state solution, under the constraint that the
total number MinD and MinE molecules matches our simulation.  We write
down as a matrix the linear response of the time derivatives of the
density to small, spatially harmonic density perturbations around the
steady-state density.  The system is stable at this wavelength
provided all eigenvalues of the matrix are negative, indicating that
all perturbations with this wavelength will decay.  We find the
stability limit by searching for the largest wavelength at which the
system is stable.  Through this method we arrive at an upper half
wavelength stability limit of $1.56\micron$ for our $0.40\micron$
flattened cells. As expected, when decreasing the lengths and widths
of our simulated flattened cells so that the longest distance across
the cell is less than this length, the cells stop exhibiting any
oscillatory behavior.  The deterministic model relaxes into a
motionless state and the stochastic model exhibits random fluctuations
without spatial oscillations.

@@

Figure~\ref{image-p} begins about 300 seconds into the simulation and
shows one period of oscillation.  At $t=0$ there is a high
concentration of MinD:ATP that has accumulated on the membrane at the
bottom of the cell. An important aspect of Huang's model is that the
MinD is attracted to and sticks to the membrane nonlinearly: as it
accumulates there it begins to `recruit' other MinD that is diffusing
in the cytoplasm nearby, causing peaks in concentration to build up on
the walls.  Meanwhile, the MinE creeps downward as it reacts with the
membrane-bound MinD, forming the MinD:MinE:ATP complex, then breaks it
apart and diffuses downward a bit more before it again reacts with
membrane-bound MinD.  This process can be seen in the form of the
well-known ``MinE rings'' (actually, MinE bound to MinD on the
membrane).  These rings are visible in the deterministic model plots
as a green band on the walls from 0 seconds to 4 seconds (and later
from 15 to 23 seconds).  The appearance of these rings in the
deterministic model along side their less obvious appearance in the
stochastic model highlights an advantage of the deterministic model:
the idealization of deterministic data allows one to see patterns in
the averaged behavior that might otherwise be missed.  In the
stochastic model the MinE still exhibits higher concentrations in the
same regions throughout the process, but what would ideally be a
``ring'' pattern is instead an asymmetric collection of maxima, which
would become a ring after phase-locked averaging.

@@

\blue{(5a)} Both Fig.~\ref{image-p} and its associated movies (see
supplementary material), show that the stochastic model exhibits a
``starry night'' effect characterized by spatially fixed points of
protein density build up, as recruitment leads to clusters of MinD
forming on the wall and then subsequently dissipating.  This results
from the lack of inclusion of diffusion along the membrane in Huang's
model, which would allow these recruitment clusters to spread out.
Experiments have confirmed that diffusion along the membrane does
occur, albeit with a rate two orders of magnitude lower than that in
the cytoplasm~\cite{meacci2005min}.  The `stars' in the effect
typically last for around 10 seconds. The membrane-bound proteins will
travel a distance of around one micron during one of these periods of
time, enough to spread out across the polar section of the one micron
diameter wild-type pill cells.  As we see below, the flattened,
irregular cells are large enough so that the one micron blur would not
completely eliminate the `starry night' variation in maxima position
effect.

@@

We also analyze the temporal periodicity of the system, which we
quantify with temporal correlation function of the total MinD found in
two opposite polar regions.  This correlation function, which is
displayed in Fig.~\ref{corr-pill}, is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom thirds of a cell.  These plots
help us in studying the periods and the regularity and stability of
oscillations.  These curves are well fit with a simple decoherence
model, given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods and in the case of the deterministic model oscillate
indefinitely with complete coherence, indicating that its behavior is
perfectly periodic.

@@

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
the flattening and accompying lateral expansion of the cell versus
cell shape irregularity and asymmetry in developing an understanding
of the origin of the experimentally observed behavior.  Although in
the movies the stadiums appear to display, on average over long time
scales, a somewhat more regular oscillation than appears in the
irregular shapes \emph{A} and \emph{B}, the limited time range of the
arrow plots (350 seconds) is inadequate to make this comparison.  We
therefore return to the correlation function between the number of
proteins in a segment at the top of the cell and the number of
proteins in a segment at the bottom of the cell.
Fig.~\ref{fig:corr-B} shows this plot for shape \emph{B} and stadium
\emph{B}.  E. Coli cell doubling times vary widely depending on
environmental conditions, but will often be between 20 minutes to 100
minutes~\cite{pierucci1972chromosome}, while fluorescent microscopy
experiments have shown that the FtsZ ring builds itself with a half
life of roughly 30 seconds~\cite{stricker2002rapid}.  Here we plot
correlations of correlation times up to 500 seconds, a biologically
relevant timescale.

We see that both in the shape \emph{B} and stadium \emph{B}, the
deterministic model is perfectly periodic.  Both the deterministic and
stochastic model correlations have the same number of proteins in
their cells, and have been normalized with the same normalization
factor, so that they can be directly compared against each other.  The
stochastic simulation of shape \emph{B} has a coherence time of 5
periods, while the corresponding simulation of stadium \emph{B} has a
coherence time of 8.6 periods.
%
The correlation function for stadium \emph{A} is similar to that of
stadium~\emph{B}, with similar coherence time (13 periods, not
shown), while shape~\emph{A} also exhibits a similar coherence time (4.9
periods).
%

From the stochastic correlation data, we conclude that irregularity of
shape has only a moderate affect on the temporal regularity of the
oscillatory behavior, with even the very distorted shape~\emph{B}
exhibiting only a factor of two greater decoherence.

>>>>

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author Summary}
  When an E. Coli cell divides it must split into two relatively equal
  volumes in order to survive.  One mechanism that aids in this
  process is a system of proteins that oscillate between cell ends and
  help the original cell divide along its center.  A recent experiment
  found that by drastically deforming these cells this oscillation can
  be disrupted.  We examine these oscillations in similarly deformed
  cells computationally, using two widely used simulation models, and
  find good agreement using one of them.  The other model fails in a
  manner far more dramatic than has been previously observed.
  Furthermore, we find that the effect observed experimentally in
  asymmetric and flattened cells is primarily a result of the
  flattening and lateral expansion, not the asymmetry.

@@

Previous experimental studies have shown that the MinD protein system
is capable of exhibiting oscillations in round
shapes~\cite{fange2006noise} as well as in connected three pronged
tube shapes~\cite{varma2008min}, in which the oscillations seem to
seek out the extreme poles in the cell.
%%   One conclusion of these
%% studies has been that the MinD system robustly forms bipolar
%% oscillations regardless of variations in cellular shape.
Mannik \emph{et al.} have recently shown that there are limitations to
this robust capability to
oscillate~\cite{mannik2012robustness,mannik2009bacterial,mannik2010bacteria}. They
have experimentally forced \emph{E. coli} cells into microfabricated
channels of thickness less than that of the natural diameter of the
cells, in which they area able to grow and divide. Upon entering the
channels, the cells undergo a mechanical deformation in which they
both widen and lengthen within the plane of the channel.  This
deformation results in very wide (they can reach widths of over
$5\micron$) flattened cells that when viewed from the top down have
irregular and asymmetric shapes.  While these cells are still able to
divide into surprisingly equal volumes, the MinD oscillations in these
cells are spatially irregular. Seen with fluorescent microscopy, the
MinD maximize in multiple locations within the cell, in a seemingly
random sequence. These experiments allow for an opportunity to test
models of the MinD system against more extreme cases than have been
seen thus far.  An additional interesting question is whether the
irregularity observed by Mannik \emph{et al.}  indicates that the MinD
system can become chaotic (as in
e.g. Ref.~\citenum{kuramoto1978diffusion}), or if this irregularity is
the result of a stochastic but nonchaotic process.
%
A number of models of the MinD protein system have been developed that
accurately describe its basic oscillatory nature.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
approach with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is
recruited to the membrane by MinD that is already clustered there
(following observed non-linear attachment of MinD on the cell
membrane~\cite{hu2002dynamic,shih2002division}), and is stationary
once attached.
%

\blue{5a,b,c,d}Several models
\cite{fange2006noise,bonny2013membrane,halatek2012highly} modify
Huang's original model to more accurately reflect experimental
findings, which were published after his 2003
paper~\cite{meacci2006mobility,loose2011min}. These models introduce
diffusion along the membrane that is two to three orders of magnitude
slower than that within the cytoplasm, and increase the rate of
cytoplasmic diffusion by a factor of about five.  They also increase
or assume instantaneous~\cite{bonny2013membrane} the nucleotide
exchange rate.  In addition, they revise the process of MinD
recruitment to the membrane, removing recruitment of MinD to the
membrane by MinD-MinE complexes, behavior that has not been observed
experimentally.  The model of Bonny \emph{et al.} further more allows
MinE to remain independently attached to the membrane after
dissasociating with MinD~\cite{bonny2013membrane}.
%

There are only on the order of 1000 MinD proteins in a given cell,
which is significantly less than the number of gridpoints used in our
simulations, suggesting that a continuum description will break down
and stochastic variation plays a significant role.  To treat this,
there have been a number of variations of the Huang 2003 model that
stochastically simulate the similar reaction-diffusion equations at
the level of individual proteins~\cite{fange2006noise,
  kerr2006division}.  These studies largely confirm the results of
Huang's deterministic model when applied to the wild-type, pill-shaped
phenotype.  However, the stochastic models are slightly more
successful in predicting experimentally observed oscillations in round
cell phenotypes~\cite{fange2006noise, huang2004min}, and they enable
prediction of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

@@

Huang's simulations use total MinD and MinE concentrations of
$1,000/\micron$ and $350/\micron$, respectively, in a cylindrical cell
of radius $0.5\micron$, and in our (non-cylindrical) cells we use the
same number of proteins per unit volume.  These concentration values
are $1273\micron^{-3}$ and $446\micron^{-3}$, respectively. We have
written our own simulation platform and membrane creation tools from
scratch, instead of using existing available software used in previous
studies~\cite{bonny2013membrane,fange2006noise,halatek2012highly,kerr2006division}.
Our simulations take place within a three-dimensional Cartesian grid
that has a grid spacing of .05\micron, and our cell shapes are
specified as the zero of an analytic function which can generate the
geometry presented in this paper.  This code is publically available
on github~\cite{git}, including scripts that generate the simulation
data used in this paper.

@@

We mean to investigate the geometric limits of the Min system
oscillations as observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}, so we have modeled the Min system
in several cell shapes and sizes.  Here we present a selection of
these, beginning with naturally occurring pill-shaped cells, followed
by a number of flattened out shapes which reflect the experiments of
Mannik \emph{et al.}, in which bacteria are confined within a thin
slits.  These slits are fabricated with a width of .25\micron, but
they are coated with a PDMS lining that the cells are able to deform,
raising uncertainty in the actual cell thickness.  Tests with pure
silicon, non-deformable slits show that the smallest thickness for
which cells are able to penetrate is
.4\micron~\cite{mannik2012robustness,mannik2009bacterial}.  We
therefore assume that the PDMS slits have been deformed to this
thickness and simulate flattened cells with a thickness of .4\micron.
Viewed from the top down the cells have the shapes described below and
viewed from the side they have at their edges a semicircular
cross-section (one may imagine the shape of a pancake).
%
In this paper we focus on four specific flattened cell shapes.  Two of
these shapes replicate those published by Mannik \emph{et al.}, and
the other two are `stadium' shapes that respectively have the same
aspect ratio, thickness, and volume as the two cell shapes
experimentally observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}.  Viewed from the top down, these
stadium shapes appear as rectangles with semi-circular end caps on the
long axis ends.

@@

Huang \emph{et al.}~\cite{huang2003dynamic} have performed a linear
stability analysis on a cylindrical model which shows an upper limit
on a steady state solution of a $2\micron$ half wavelength.  Cells
with dimensions longer than this spontaneously develop spatial
oscillations, while cells that are shorter relax into a homogeneous
steady state.  We have performed a similar linear stability analysis
on an infinite slab with a thickness equal to that of our flattened
cells ($0.40\micron$).  We first solve the model's five differential
equations for the steady state solution, under the constraint that the
total number MinD and MinE molecules matches our simulation.  We write
down as a matrix the linear response of the time derivatives of the
density to small, spatially harmonic density perturbations around the
steady-state density.  The system is stable at this wavelength
provided all eigenvalues of the matrix are negative, indicating that
all perturbations with this wavelength will decay.  We find the
stability limit by searching for the largest wavelength at which the
system is stable.  Through this method we arrive at an upper half
wavelength stability limit of $1.56\micron$ for our $0.40\micron$
flattened cells. As expected, when decreasing the lengths and widths
of our simulated flattened cells so that the longest distance across
the cell is less than this length, the cells stop exhibiting any
oscillatory behavior.  The deterministic model relaxes into a
motionless state and the stochastic model exhibits random fluctuations
without spatial oscillations.

@@

Figure~\ref{image-p} begins about 300 seconds into the simulation and
shows one period of oscillation.  At $t=0$ there is a high
concentration of MinD:ATP that has accumulated on the membrane at the
bottom of the cell. An important aspect of Huang's model is that the
MinD is attracted to and sticks to the membrane nonlinearly: as it
accumulates there it begins to `recruit' other MinD that is diffusing
in the cytoplasm nearby, causing peaks in concentration to build up on
the walls.  Meanwhile, the MinE creeps downward as it reacts with the
membrane-bound MinD, forming the MinD:MinE:ATP complex, then breaks it
apart and diffuses downward a bit more before it again reacts with
membrane-bound MinD.  This process can be seen in the form of the
well-known ``MinE rings'' (actually, MinE bound to MinD on the
membrane).  These rings are visible in the deterministic model plots
as a green band on the walls from 0 seconds to 4 seconds (and later
from 15 to 23 seconds).  The appearance of these rings in the
deterministic model alongside their less obvious appearance in the
stochastic model highlights an advantage of the deterministic model:
the idealization of deterministic data allows one to see patterns in
the averaged behavior that might otherwise be missed.  In the
stochastic model the MinE still exhibits higher concentrations in the
same regions throughout the process, but what would ideally be a
``ring'' pattern is instead an asymmetric collection of maxima, which
would become a ring after phase-locked averaging.

@@

\blue{(5a)} Both Fig.~\ref{image-p} and its associated movies (see
supplementary material), show that the stochastic model exhibits a
``starry night'' effect characterized by spatially fixed points of
protein density build up, as recruitment leads to clusters of MinD
forming on the wall and then subsequently dissipating.  This results
from the omission of diffusion along the membrane from Huang's model,
which would allow these recruitment clusters to spread out.
Experiments have confirmed that diffusion along the membrane does
occur, albeit with a rate two orders of magnitude lower than that in
the cytoplasm~\cite{meacci2005min}.  The `stars' in the effect
typically last for around 10 seconds. The membrane-bound proteins will
travel a distance of around one micron during one of these periods of
time, enough to spread out across the polar section of the one micron
diameter wild-type pill cells.  As we see below, the flattened,
irregular cells are large enough so that the one micron blurring would
not completely eliminate the `starry night' variation in maxima
position effect.

@@

We also analyze the temporal periodicity of the system, which we
quantify with the temporal correlation function of the total MinD
found in two opposite polar regions.  This correlation function, which
is displayed in Fig.~\ref{corr-pill}, is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom thirds of a cell.  These plots
help us in studying the periods and the regularity and stability of
oscillations.  These curves are well fit with a simple decoherence
model, given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods and in the case of the deterministic model oscillate
indefinitely with complete coherence, indicating that its behavior is
periodic.

@@

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
the flattening and accompying lateral expansion of the cell versus
cell shape irregularity and asymmetry in the temporal regularity of
the oscillations.  Although in the movies the stadiums appear to
display, on average over long time scales, a somewhat more regular
oscillation than appears in the irregular shapes \emph{A} and
\emph{B}, the limited time range of the arrow plots (350 seconds) is
inadequate to make this comparison.  We therefore return to the
correlation function between the number of proteins in a segment at
the top of the cell and the number of proteins in a segment at the
bottom of the cell.  Fig.~\ref{fig:corr-B} shows this plot for shape
\emph{B} and stadium \emph{B}.  Here we plot correlations of
correlation times up to 500 seconds, a biologically relevant
timescale.  E. Coli cell doubling times vary widely depending on
environmental conditions, but will often be between 20 minutes to 100
minutes~\cite{pierucci1972chromosome}, while fluorescent microscopy
experiments have shown that the FtsZ ring builds itself with a half
life of roughly 30 seconds~\cite{stricker2002rapid}.

We see that both in the shape \emph{B} and stadium \emph{B}, the
deterministic model is perfectly periodic.  Both the deterministic and
stochastic model correlations have the same number of proteins in
their cells, and have been normalized with the same normalization
factor, so that they can be directly compared against each other.  The
stochastic simulation of shape \emph{B} has a coherence time of 5
periods, while the corresponding simulation of stadium \emph{B} has a
coherence time of 8.6 periods.
%
The correlation function for stadium \emph{A} is similar to that of
stadium~\emph{B}, with similar coherence time (13 periods, not shown),
while shape~\emph{A} also exhibits a similar coherence time (4.9
periods).  This confirms that the stadium shapes do indeed oscillate
somewhat more regularly than the irrelgular shapes, but both are
comparable to the wild-type pill shape.
%
From the stochastic correlation data, we conclude that irregularity of
shape has only a moderate affect on the temporal regularity of the
oscillatory behavior, with even the very distorted shape~\emph{B}
exhibiting only a factor of two greater decoherence.
